---
title: "Predicting Ship Carbon Intensity Rating: A Machine Learning Classification Approach"
author: "Shipping Performance Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  fig.align  = "center"
)

# PNG export directory
plot_dir <- "plots"
if (!dir.exists(plot_dir)) dir.create(plot_dir)
save_plot <- function(name, width = 10, height = 6) {
  ggsave(file.path(plot_dir, paste0(name, ".png")),
         plot = last_plot(), width = width, height = height, dpi = 150)
}
```

---

# Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(caret)
library(ranger)
library(gbm)
library(nnet)
library(kknn)
library(pROC)
library(knitr)
library(kableExtra)
library(scales)
library(viridis)
library(gridExtra)
library(RColorBrewer)
```

---

# 1. Problem Statement

**Objective:** Predict a vessel's **Carbon Intensity Indicator (CII) Rating** (A / B / C / D / E)
from its operational and vessel characteristics — a 5-class classification problem.

**Why this target?**
The CII Rating is an IMO (International Maritime Organization) regulatory metric that grades
vessels on carbon efficiency. It directly impacts compliance, port fees, and charterer
decisions. Predicting the CII class from operational parameters (speed, load, engine power,
capacity, fuel type) enables fleet managers to plan voyages that achieve better ratings.

> The previous classifier used `Ship_Performance_Dataset.csv` with a synthetic
> `Efficiency_Class` target — diagnostic testing confirmed all feature-target correlations
> were < 0.02 (pure noise). This version uses the **Modernized dataset** which has
> genuine physical relationships (|r| up to 0.92 between columns).

**Target variable:** `CII_Rating` — assigned by the IMO CII calculation framework:

| Rating | Description | Implication |
|--------|-------------|-------------|
| A | Superior — significantly below required CII | Best — green rating |
| B | Minor major — below required CII | Good |
| C | Moderate — within required CII limit | Acceptable |
| D | Minor deficiency — above required CII | Requires improvement plan |
| E | Major deficiency — significantly above required CII | Mandatory corrective action |

> **Leakage prevention:** `CII_Value` (the raw score from which the rating is derived),
> `CO2_Emissions_tons`, and `Emission_Factor` are **excluded** from the feature set —
> they are direct inputs to the CII calculation and would trivially leak the answer.

**Business value:** Early prediction of CII Rating before a voyage allows operators to
adjust speed, fuel type, and load strategy to avoid D/E ratings and stay compliant.

---

# 2. Data Loading & Preprocessing

## 2.1 Load Data

```{r load}
df <- read.csv("../data/Ship_Performance_Dataset_Modernized.csv",
               stringsAsFactors = FALSE)
df$Date <- as.Date(df$Date)

cat("Dimensions:", nrow(df), "rows ×", ncol(df), "cols\n")
cat("\nColumns:\n")
cat(paste(" •", names(df), collapse = "\n"), "\n")
```

## 2.2 Target Distribution

```{r target-dist}
# Treat empty Maintenance_Status as NA (136 rows) — will be removed at drop_na()
df$Maintenance_Status[df$Maintenance_Status == ""] <- NA

# Make CII_Rating an ordered factor A > B > C > D > E (A = best)
df$CII_Rating <- factor(df$CII_Rating, levels = c("A","B","C","D","E"), ordered = TRUE)

cat("CII_Rating distribution:\n")
print(table(df$CII_Rating))
cat("\nProportions (%):\n")
print(round(prop.table(table(df$CII_Rating)) * 100, 1))
```

```{r class-dist-plot, fig.height=4}
cii_palette <- c("A" = "#27ae60", "B" = "#2ecc71",
                 "C" = "#f39c12", "D" = "#e67e22", "E" = "#e74c3c")

class_df <- as.data.frame(table(df$CII_Rating)) %>%
  rename(Rating = Var1, Count = Freq) %>%
  mutate(Pct = round(Count / sum(Count) * 100, 1))

ggplot(class_df, aes(x = Rating, y = Count, fill = Rating)) +
  geom_col(color = "white", width = 0.65) +
  geom_text(aes(label = paste0(Count, "\n(", Pct, "%)")),
            vjust = -0.3, fontface = "bold", size = 4.5) +
  scale_fill_manual(values = cii_palette) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Target Class Distribution — CII Rating (A = Best, E = Worst)",
       subtitle = "Near-perfectly balanced across 5 classes — 20% random baseline",
       x = "CII Rating", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

save_plot("01_target_distribution", height = 4)
```

## 2.3 Feature–Target Signal Check

Verify real signal exists before modelling using ANOVA F-statistics.

```{r signal-check, fig.height=5}
num_cols <- c("Speed_Over_Ground_knots", "Engine_Power_kW", "Distance_Traveled_nm",
              "Draft_meters", "Cargo_Weight_tons", "Operational_Cost_USD",
              "Revenue_per_Voyage_USD", "Turnaround_Time_hours",
              "Efficiency_nm_per_kWh", "Seasonal_Impact_Score",
              "Weekly_Voyage_Count", "Average_Load_Percentage",
              "Estimated_Capacity_DWT")

fstats <- sapply(num_cols, function(col) {
  m <- aov(as.formula(paste(col, "~ CII_Rating")), data = df)
  summary(m)[[1]][["F value"]][1]
})

fstat_df <- data.frame(
  Feature = num_cols,
  F_Stat  = round(fstats, 1),
  Signal  = ifelse(fstats > 100, "Very Strong",
             ifelse(fstats > 10,  "Strong",
             ifelse(fstats > 3,   "Moderate", "Weak")))
)

ggplot(fstat_df, aes(x = reorder(Feature, F_Stat), y = F_Stat, fill = Signal)) +
  geom_col(color = "white", width = 0.7) +
  geom_hline(yintercept = 3,  linetype = "dashed", color = "#e74c3c",  linewidth = 0.8) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "#f39c12", linewidth = 0.8) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "#27ae60", linewidth = 0.8) +
  annotate("text", x = 1.5, y = 6,   label = "Moderate (F=3)",  color = "#e74c3c", size = 3.5) +
  annotate("text", x = 1.5, y = 15,  label = "Strong (F=10)",   color = "#f39c12", size = 3.5) +
  annotate("text", x = 1.5, y = 110, label = "Very Strong (F=100)", color = "#27ae60", size = 3.5) +
  scale_fill_manual(values = c("Very Strong" = "#27ae60", "Strong" = "#2ecc71",
                                "Moderate"    = "#f39c12", "Weak"   = "#e74c3c")) +
  coord_flip() +
  labs(title = "ANOVA F-Statistics — Feature Signal Strength for CII_Rating",
       subtitle = "Very Strong signal detected in Estimated_Capacity_DWT, Cargo_Weight_tons, Engine_Power_kW",
       x = NULL, y = "F-Statistic (log scale note: axis not logged)", fill = "Signal") +
  theme_minimal(base_size = 12)

save_plot("02_feature_signal_anova", height = 5)

kable(fstat_df[order(-fstat_df$F_Stat), ],
      caption = "Feature Signal Strength — ANOVA F vs CII_Rating",
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  column_spec(3, bold = TRUE,
              color = ifelse(fstat_df[order(-fstat_df$F_Stat),]$Signal %in% c("Very Strong","Strong"),
                             "darkgreen", ifelse(
                             fstat_df[order(-fstat_df$F_Stat),]$Signal == "Moderate",
                             "darkorange", "red")))
```

## 2.4 Build Feature Matrix

```{r preprocess}
# Exclude: Date, CII_Value (direct leakage), CO2_Emissions_tons (leakage),
#          Emission_Factor (leakage)
# Keep: all operational features + Fuel_Type + Estimated_Capacity_DWT
df_model <- df %>%
  dplyr::select(-Date, -CII_Value, -CO2_Emissions_tons, -Emission_Factor) %>%
  mutate(
    Ship_Type          = factor(Ship_Type),
    Route_Type         = factor(Route_Type),
    Engine_Type        = factor(Engine_Type),
    Weather_Condition  = factor(Weather_Condition),
    Maintenance_Status = factor(Maintenance_Status),
    Fuel_Type          = factor(Fuel_Type)
  ) %>%
  drop_na()

cat("Model dataset:", nrow(df_model), "rows ×", ncol(df_model), "cols\n")
cat("Rows dropped (NA):", nrow(df) - nrow(df_model), "\n")
cat("\nTarget distribution after cleaning:\n")
print(table(df_model$CII_Rating))
```

```{r feature-list}
feature_cols <- setdiff(names(df_model), "CII_Rating")
cat("Features (", length(feature_cols), "):\n")
cat(paste(" •", feature_cols, collapse = "\n"), "\n")
```

---

# 3. Train / Test Split

**Strategy:** Stratified 80/20 split — preserves class proportions in both sets.

```{r split}
set.seed(42)
train_idx <- createDataPartition(df_model$CII_Rating, p = 0.80, list = FALSE)

train_df <- df_model[ train_idx, ]
test_df  <- df_model[-train_idx, ]

cat("Training set:", nrow(train_df), "rows\n")
cat("Test set:    ", nrow(test_df),  "rows\n\n")

cat("Train class proportions (%):\n")
print(round(prop.table(table(train_df$CII_Rating)) * 100, 1))
cat("\nTest class proportions (%):\n")
print(round(prop.table(table(test_df$CII_Rating)) * 100, 1))
```

---

# 4. Cross-Validation Setup

**Strategy:** 5-fold stratified cross-validation on the training set.
All models share the same CV folds for a fair comparison.

```{r cv-setup}
cv_ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final",
  verboseIter     = FALSE
)

cat("CV: 5-fold stratified | Metric: Accuracy + Macro F1\n")
cat("Baseline (random chance 5 classes): 20.0%\n")
```

---

# 5. Algorithm 1 — Multinomial Logistic Regression

**Baseline linear model.** Extends binary logistic regression to 5 classes using
a softmax function — draws linear decision boundaries in feature space.

```{r lr-train}
set.seed(42)
lr_model <- train(
  CII_Rating ~ .,
  data       = train_df,
  method     = "multinom",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  trace      = FALSE
)

cat("Logistic Regression — Best CV Accuracy:",
    round(max(lr_model$results$Accuracy), 4), "\n")
```

```{r lr-predict}
lr_pred  <- predict(lr_model, test_df)
lr_probs <- predict(lr_model, test_df, type = "prob")

lr_cm <- confusionMatrix(lr_pred, test_df$CII_Rating)
cat("=== Logistic Regression — Test Set ===\n")
print(lr_cm$overall[c("Accuracy", "Kappa")])
cat("\nPer-class Precision / Recall / F1:\n")
print(round(lr_cm$byClass[, c("Precision", "Recall", "F1")], 4))
```

---

# 6. Algorithm 2 — K-Nearest Neighbours (KNN)

**Instance-based learner.** No explicit training — classifies each test point by
majority vote among its K closest neighbours in the scaled feature space.

```{r knn-train}
set.seed(42)
knn_model <- train(
  CII_Rating ~ .,
  data       = train_df,
  method     = "kknn",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  tuneGrid   = expand.grid(
    kmax     = c(5, 7, 9, 11, 15),
    distance = 2,
    kernel   = "optimal"
  )
)

cat("KNN — Best K:", knn_model$bestTune$kmax, "\n")
cat("KNN — Best CV Accuracy:", round(max(knn_model$results$Accuracy), 4), "\n")
```

```{r knn-tune-plot, fig.height=4}
plot(knn_model, main = "KNN — Accuracy vs K (5-fold CV)")
```

```{r knn-predict}
knn_pred  <- predict(knn_model, test_df)
knn_probs <- predict(knn_model, test_df, type = "prob")

knn_cm <- confusionMatrix(knn_pred, test_df$CII_Rating)
cat("=== KNN — Test Set ===\n")
print(knn_cm$overall[c("Accuracy", "Kappa")])
cat("\nPer-class Precision / Recall / F1:\n")
print(round(knn_cm$byClass[, c("Precision", "Recall", "F1")], 4))
```

---

# 7. Algorithm 3 — Random Forest

**Ensemble of 500 decision trees** using bootstrap aggregation and random feature
subsets at each split — reduces variance and handles non-linear relationships well.

```{r rf-train}
set.seed(42)
rf_model <- train(
  CII_Rating ~ .,
  data       = train_df,
  method     = "ranger",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  importance = "impurity",
  tuneGrid   = expand.grid(
    mtry          = c(3, 5, 7, 9),
    splitrule     = "gini",
    min.node.size = c(1, 5)
  )
)

cat("Random Forest — Best mtry:", rf_model$bestTune$mtry, "\n")
cat("Random Forest — Best min.node.size:", rf_model$bestTune$min.node.size, "\n")
cat("Random Forest — Best CV Accuracy:", round(max(rf_model$results$Accuracy), 4), "\n")
```

```{r rf-predict}
rf_pred  <- predict(rf_model, test_df)
rf_probs <- predict(rf_model, test_df, type = "prob")

rf_cm <- confusionMatrix(rf_pred, test_df$CII_Rating)
cat("=== Random Forest — Test Set ===\n")
print(rf_cm$overall[c("Accuracy", "Kappa")])
cat("\nPer-class Precision / Recall / F1:\n")
print(round(rf_cm$byClass[, c("Precision", "Recall", "F1")], 4))
```

```{r rf-importance, fig.height=6}
plot(varImp(rf_model), top = 15,
     main = "Random Forest — Top 15 Feature Importances")
```

---

# 8. Algorithm 4 — Gradient Boosting Machine (GBM)

**Sequential boosting ensemble.** Each tree corrects the residual errors of all prior
trees via gradient descent — typically the strongest performer on structured tabular data.

```{r gbm-train}
set.seed(42)
gbm_model <- train(
  CII_Rating ~ .,
  data       = train_df,
  method     = "gbm",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  verbose    = FALSE,
  tuneGrid   = expand.grid(
    n.trees           = c(100, 200),
    interaction.depth = c(3, 5),
    shrinkage         = c(0.05, 0.1),
    n.minobsinnode    = 10
  )
)

cat("GBM — Best n.trees:", gbm_model$bestTune$n.trees, "\n")
cat("GBM — Best interaction.depth:", gbm_model$bestTune$interaction.depth, "\n")
cat("GBM — Best shrinkage:", gbm_model$bestTune$shrinkage, "\n")
cat("GBM — Best CV Accuracy:", round(max(gbm_model$results$Accuracy), 4), "\n")
```

```{r gbm-predict}
gbm_pred  <- predict(gbm_model, test_df)
gbm_probs <- predict(gbm_model, test_df, type = "prob")

gbm_cm <- confusionMatrix(gbm_pred, test_df$CII_Rating)
cat("=== GBM — Test Set ===\n")
print(gbm_cm$overall[c("Accuracy", "Kappa")])
cat("\nPer-class Precision / Recall / F1:\n")
print(round(gbm_cm$byClass[, c("Precision", "Recall", "F1")], 4))
```

```{r gbm-importance, fig.height=6}
plot(varImp(gbm_model), top = 15,
     main = "GBM — Top 15 Feature Importances")
```

---

# 9. Model Comparison

## 9.1 Macro F1-Score Table

```{r macro-f1}
macro_f1 <- function(cm) round(mean(cm$byClass[, "F1"], na.rm = TRUE), 4)

results_df <- data.frame(
  Algorithm     = c("Logistic Regression", "KNN", "Random Forest", "GBM"),
  CV_Accuracy   = round(c(max(lr_model$results$Accuracy),
                           max(knn_model$results$Accuracy),
                           max(rf_model$results$Accuracy),
                           max(gbm_model$results$Accuracy)), 4),
  Test_Accuracy = round(c(lr_cm$overall["Accuracy"],
                           knn_cm$overall["Accuracy"],
                           rf_cm$overall["Accuracy"],
                           gbm_cm$overall["Accuracy"]), 4),
  Kappa         = round(c(lr_cm$overall["Kappa"],
                           knn_cm$overall["Kappa"],
                           rf_cm$overall["Kappa"],
                           gbm_cm$overall["Kappa"]), 4),
  Macro_F1      = c(macro_f1(lr_cm), macro_f1(knn_cm),
                    macro_f1(rf_cm), macro_f1(gbm_cm))
)

best_row <- which.max(results_df$Macro_F1)

kable(results_df, caption = "Model Comparison — All Metrics (Baseline = 20.0%)") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  row_spec(best_row, bold = TRUE, background = "#d5f5e3") %>%
  column_spec(5, bold = TRUE)
```

## 9.2 Accuracy Comparison Plot

```{r comparison-plot, fig.height=5}
results_long <- results_df %>%
  pivot_longer(cols = c(CV_Accuracy, Test_Accuracy, Macro_F1),
               names_to = "Metric", values_to = "Score")

ggplot(results_long, aes(x = reorder(Algorithm, Score), y = Score, fill = Metric)) +
  geom_col(position = "dodge", color = "white", width = 0.7) +
  geom_hline(yintercept = 0.20, linetype = "dashed",
             color = "#e74c3c", linewidth = 0.8) +
  annotate("text", x = 0.7, y = 0.215, label = "Random baseline (20.0%)",
           color = "#e74c3c", size = 3.5) +
  geom_text(aes(label = round(Score, 3)),
            position = position_dodge(width = 0.7),
            hjust = -0.1, size = 3, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 1.1), labels = percent_format()) +
  coord_flip() +
  labs(title = "Model Comparison — CV Accuracy vs Test Accuracy vs Macro F1",
       subtitle = "Baseline = 20% (5-class random chance)",
       x = NULL, y = "Score", fill = "Metric") +
  theme_minimal(base_size = 13)

save_plot("03_model_comparison", height = 5)
```

## 9.3 CV Resampling Distribution

```{r cv-resample-plot, fig.height=5}
resamps <- resamples(list(LR  = lr_model, KNN = knn_model,
                          RF  = rf_model, GBM = gbm_model))

bwplot(resamps, metric = "Accuracy",
       main = "5-Fold CV Accuracy Distribution — All Models")
```

---

# 10. Confusion Matrices

```{r confusion-matrices, fig.height=12}
plot_cm <- function(cm_obj, title, palette) {
  cm_tbl <- as.data.frame(cm_obj$table)
  names(cm_tbl) <- c("Predicted", "Actual", "Freq")
  ggplot(cm_tbl, aes(x = Actual, y = Predicted, fill = Freq)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = Freq), fontface = "bold", size = 4.5) +
    scale_fill_gradient(low = "white", high = palette) +
    labs(title = title, x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")
}

p_lr  <- plot_cm(lr_cm,  "Logistic Regression", "#3498db")
p_knn <- plot_cm(knn_cm, "KNN",                 "#2ecc71")
p_rf  <- plot_cm(rf_cm,  "Random Forest",        "#9b59b6")
p_gbm <- plot_cm(gbm_cm, "GBM",                 "#e67e22")

grid.arrange(p_lr, p_knn, p_rf, p_gbm, ncol = 2,
             top = "Confusion Matrices — Test Set (Rows = Predicted, Cols = Actual)")
```

```{r save-cm, include=FALSE}
cm_grid <- arrangeGrob(p_lr, p_knn, p_rf, p_gbm, ncol = 2)
ggsave(file.path(plot_dir, "04_confusion_matrices.png"),
       cm_grid, width = 12, height = 12, dpi = 150)
```

---

# 11. Per-Class Metrics

```{r per-class-metrics, fig.height=8}
extract_metrics <- function(cm, algo_name) {
  as.data.frame(cm$byClass[, c("Precision","Recall","F1")]) %>%
    rownames_to_column("Class") %>%
    mutate(Class     = str_remove(Class, "Class: "),
           Algorithm = algo_name)
}

per_class <- bind_rows(
  extract_metrics(lr_cm,  "LR"),
  extract_metrics(knn_cm, "KNN"),
  extract_metrics(rf_cm,  "RF"),
  extract_metrics(gbm_cm, "GBM")
) %>%
  pivot_longer(cols = c(Precision, Recall, F1),
               names_to = "Metric", values_to = "Value")

ggplot(per_class, aes(x = Algorithm, y = Value, fill = Algorithm)) +
  geom_col(position = "dodge", color = "white") +
  geom_text(aes(label = round(Value, 2)),
            vjust = -0.3, size = 2.8, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 1.2), labels = percent_format()) +
  facet_grid(Class ~ Metric) +
  labs(title = "Per-Class Precision / Recall / F1 by Algorithm",
       x = NULL, y = "Score") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")

save_plot("05_per_class_metrics", height = 8)
```

---

# 12. ROC Curves (One-vs-Rest)

```{r roc-curves, fig.height=10}
classes <- levels(test_df$CII_Rating)

par(mfrow = c(3, 2), mar = c(4, 4, 3, 1))

for (cls in classes) {
  actual_bin <- as.numeric(test_df$CII_Rating == cls)
  roc_lr  <- roc(actual_bin, lr_probs[[cls]],  quiet = TRUE)
  roc_knn <- roc(actual_bin, knn_probs[[cls]], quiet = TRUE)
  roc_rf  <- roc(actual_bin, rf_probs[[cls]],  quiet = TRUE)
  roc_gbm <- roc(actual_bin, gbm_probs[[cls]], quiet = TRUE)

  plot(roc_lr, col = "#3498db", lwd = 2,
       main = paste("ROC — Rating", cls, "vs Rest"),
       xlab = "False Positive Rate", ylab = "True Positive Rate")
  lines(roc_knn, col = "#2ecc71", lwd = 2)
  lines(roc_rf,  col = "#9b59b6", lwd = 2)
  lines(roc_gbm, col = "#e67e22", lwd = 2)
  abline(0, 1, lty = 2, col = "grey60")
  legend("bottomright",
         legend = c(paste("LR  AUC =", round(auc(roc_lr),  3)),
                    paste("KNN AUC =", round(auc(roc_knn), 3)),
                    paste("RF  AUC =", round(auc(roc_rf),  3)),
                    paste("GBM AUC =", round(auc(roc_gbm), 3))),
         col = c("#3498db","#2ecc71","#9b59b6","#e67e22"),
         lwd = 2, cex = 0.75, bty = "n")
}

par(mfrow = c(1,1))
```

```{r auc-table}
auc_df <- data.frame(
  Rating = classes,
  LR  = sapply(classes, function(c) round(auc(roc(as.numeric(test_df$CII_Rating==c), lr_probs[[c]],  quiet=TRUE)), 4)),
  KNN = sapply(classes, function(c) round(auc(roc(as.numeric(test_df$CII_Rating==c), knn_probs[[c]], quiet=TRUE)), 4)),
  RF  = sapply(classes, function(c) round(auc(roc(as.numeric(test_df$CII_Rating==c), rf_probs[[c]],  quiet=TRUE)), 4)),
  GBM = sapply(classes, function(c) round(auc(roc(as.numeric(test_df$CII_Rating==c), gbm_probs[[c]], quiet=TRUE)), 4))
)

kable(auc_df, caption = "AUC per Class (One-vs-Rest) — All Models") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  column_spec(which.max(colMeans(auc_df[,-1])) + 1, bold = TRUE, background = "#d5f5e3")
```

---

# 13. Feature Importance Comparison

```{r feature-importance-compare, fig.height=7}
rf_imp_df <- varImp(rf_model)$importance %>%
  rownames_to_column("Feature") %>%
  mutate(Importance = rowMeans(dplyr::select(., -Feature)),
         Algorithm  = "Random Forest") %>%
  dplyr::select(Feature, Importance, Algorithm) %>%
  slice_max(Importance, n = 10)

gbm_imp_df <- varImp(gbm_model)$importance %>%
  rownames_to_column("Feature") %>%
  mutate(Importance = rowMeans(dplyr::select(., -Feature)),
         Algorithm  = "GBM") %>%
  dplyr::select(Feature, Importance, Algorithm) %>%
  slice_max(Importance, n = 10)

imp_combined <- bind_rows(rf_imp_df, gbm_imp_df)

ggplot(imp_combined,
       aes(x = reorder(Feature, Importance), y = Importance, fill = Algorithm)) +
  geom_col(color = "white", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Random Forest" = "#9b59b6", "GBM" = "#e67e22")) +
  facet_wrap(~ Algorithm, scales = "free_x") +
  labs(title = "Top 10 Feature Importances — Random Forest vs GBM",
       subtitle = "Features that drive CII Rating prediction",
       x = NULL, y = "Importance Score") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

save_plot("06_feature_importance", height = 7)
```

---

# 14. Best Model Deep Dive

```{r best-model-id}
best_algo <- results_df$Algorithm[which.max(results_df$Macro_F1)]
best_f1   <- max(results_df$Macro_F1)
best_acc  <- results_df$Test_Accuracy[which.max(results_df$Macro_F1)]
cat("Best model:", best_algo, "\n")
cat("Macro F1:", best_f1, "| Test Accuracy:", best_acc, "\n")
cat("Improvement over random baseline (20.0%):",
    round((best_acc - 0.20) * 100, 1), "percentage points\n")
```

## 14.1 Detailed Classification Report

```{r best-model-report}
best_cm <- switch(best_algo,
  "Logistic Regression" = lr_cm,
  "KNN"                 = knn_cm,
  "Random Forest"       = rf_cm,
  "GBM"                 = gbm_cm
)

detail_df <- as.data.frame(best_cm$byClass) %>%
  rownames_to_column("Class") %>%
  mutate(Class = str_remove(Class, "Class: ")) %>%
  rename(Balanced_Accuracy = `Balanced Accuracy`) %>%
  dplyr::select(Class, Precision, Recall, F1, Specificity, Balanced_Accuracy) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

kable(detail_df, caption = paste("Detailed Classification Report —", best_algo)) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  column_spec(4, bold = TRUE, background = "#eaf4fb")
```

## 14.2 Prediction Confidence Distribution

```{r confidence-dist, fig.height=5}
best_probs <- switch(best_algo,
  "Logistic Regression" = lr_probs,
  "KNN"                 = knn_probs,
  "Random Forest"       = rf_probs,
  "GBM"                 = gbm_probs
)
best_pred <- switch(best_algo,
  "Logistic Regression" = lr_pred,
  "KNN"                 = knn_pred,
  "Random Forest"       = rf_pred,
  "GBM"                 = gbm_pred
)

conf_df <- best_probs %>%
  mutate(Predicted = best_pred,
         Actual    = test_df$CII_Rating,
         Correct   = Predicted == Actual,
         Max_Prob  = pmax(A, B, C, D, E))

ggplot(conf_df, aes(x = Max_Prob, fill = Correct)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.85) +
  scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                    labels = c("TRUE" = "Correct", "FALSE" = "Wrong")) +
  facet_wrap(~ Actual, ncol = 5) +
  labs(title = paste(best_algo, "— Prediction Confidence by Actual CII Rating"),
       subtitle = "Green = correct | Red = wrong | X-axis = max predicted class probability",
       x = "Max Class Probability", y = "Count", fill = "Prediction") +
  theme_minimal(base_size = 12)

save_plot("07_confidence_distribution", height = 5)
```

## 14.3 CII Rating Boundary Errors

```{r boundary-errors, fig.height=5}
# Which ratings are most often confused? Show error heatmap for best model
err_df <- as.data.frame(best_cm$table)
names(err_df) <- c("Predicted", "Actual", "Count")

ggplot(err_df, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Count), fontface = "bold", size = 6) +
  scale_fill_gradient2(low = "white", mid = "#f39c12", high = "#e74c3c",
                       midpoint = max(err_df$Count[err_df$Actual != err_df$Predicted]) / 2) +
  labs(title = paste("Confusion Heatmap —", best_algo, "(Best Model)"),
       subtitle = "Off-diagonal = errors | Adjacent-class errors (e.g. A→B) are less harmful",
       x = "Actual CII Rating", y = "Predicted CII Rating") +
  theme_minimal(base_size = 13)

save_plot("08_best_model_confusion", height = 5)
```

---

# 15. Final Summary

```{r final-summary, echo=FALSE}
summary_tbl <- data.frame(
  Item = c(
    "Dataset",
    "Previous dataset (rejected)",
    "Target variable",
    "Why CII_Rating?",
    "Leakage exclusions",
    "New features used",
    "Classes",
    "Class balance",
    "Rows after cleaning",
    "Train / Test split",
    "Cross-validation",
    "Feature scaling",
    "Algorithms tried",
    "Primary metric",
    "Best algorithm",
    "Best Macro F1",
    "Best Test Accuracy",
    "Improvement over baseline"
  ),
  Detail = c(
    "Ship_Performance_Dataset_Modernized.csv — 2,736 rows × 24 columns",
    "Ship_Performance_Dataset.csv — all feature-target correlations < 0.02 (pure noise)",
    "CII_Rating (A / B / C / D / E) — IMO Carbon Intensity Indicator",
    "Real physical signal: Estimated_Capacity_DWT r=-0.64, Cargo r=-0.61, Engine_Power r=+0.46",
    "CII_Value, CO2_Emissions_tons, Emission_Factor (all used in CII calculation)",
    "Fuel_Type, Estimated_Capacity_DWT — new in Modernized dataset",
    "5 ordered classes: A < B < C < D < E (A = best, E = worst)",
    "~20% per class — near-perfectly balanced; no SMOTE required",
    paste0(nrow(df_model), " rows (", nrow(df) - nrow(df_model), " dropped — empty Maintenance_Status)"),
    "80% train / 20% test — stratified by CII_Rating class",
    "5-fold stratified CV on training set",
    "Z-score (centre + scale) inside caret pipeline",
    "Logistic Regression, KNN, Random Forest, GBM",
    "Macro F1-Score (equal weight across all 5 classes)",
    best_algo,
    as.character(best_f1),
    as.character(best_acc),
    paste0(round((best_acc - 0.20) * 100, 1), " percentage points above 20.0% baseline")
  )
)

kable(summary_tbl, caption = "Classification Pipeline — Final Summary") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  row_spec(c(15,16,17,18), bold = TRUE, background = "#d5f5e3")
```

---

*End of Classification Report — Predicting Ship Carbon Intensity Rating*
